### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: clean-up some notes.
Desc@ICFP: merge Colors and Spaces.

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}. The reader will notice
that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer. For the same reason, we introduce
$\DProd{\!}{\!}$ that overlaps with $\DSigma{\!}{\!}$.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where                                      \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                    \\
    \DConst{}       & \Set \To \Desc                           \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{\V{X}}          &  \V{X}                                           \\
\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
\end{array}
}
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

\[\stk{
\VecD : \Set \To \TagIDesc{\Nat} \\
\begin{array}{@{}ll}
\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
            \begin{array}{l}
                \pair{[]}{\_}{} 
                \\\\
                \bigRedBracket{
                \begin{array}{l} 
                  \LAM{n} \caseB{n}{\NatZero \To [\VecNil] \\
                                   (\NatSuc n) \To [\VecCons{}{}]} \\
                  \LAM{n} \caseB{n}{\NatZero \To [\DConst{Unit}] \\
                                   (\NatSuc n) \To [\DSigma{X}{\DVar{n}}]}
                \end{array}}
            \end{array}}
\end{array}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

@@

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
\end{array}
}
\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
                          \end{array}
                          }
\end{array}
}\]

Intuitively, the constraint is captured as follows. First, we store an
element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
stipulates that $m$ cannot be \emph{any} natural numbers: it must be
``the index minus one''. This translates into the constraint $n
\PropEq \NatSuc{m}$.

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{n}
                                         \end{array}
                                         }
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PITEL{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{(E,F)}{I}{X} \mapsto
    \pair{\pair{\ListCons{\DVar{}}{(\fst{E})}} 
               {\LAM{i}
                \pair{(\DConst{(X\: i)})}
                     {((\snd{E})\: i)}{}}{}}
         {F}{}
}\]

@@

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
                                              \DProd{}{},
                                              \DSigma{}{}, 
                                              \DPi{}{} \red{]}) \\
                                  & \bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}
\end{array}
}\]

@@

\note{ Should we stop here? Or show off some madness? }

>>>>

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}. The reader will notice
that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is
interpreted as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally,
$\DConst{\V{X}}$ and $\DSigma{\V{X}}{\DUnit}$ are equivalent. However,
$\DConst{\!}$ is more succinct. More importantly, $\DConst$ is
\emph{first-order}, unlike its equivalent encoding. From a
definitional perspective, we are giving more opportunities to the
type-system, hence reducing the burden on the programmer. For the same
reason, we introduce $\DProd{\!}{\!}$ that overlaps with
$\DSigma{\!}{\!}$.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where                                      \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                    \\
    \DConst{\!}     & \Set \To \Desc                           \\
    \DProd{\!}{\!}  & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
    \DSigma{\!}{\!} & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
    \DPi{\!}{\!}    & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{\V{X}}          &  \V{X}                                           \\
\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
\end{array}
}
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

\[\stk{
\VecD : \Set \To \TagIDesc{\Nat} \\
\begin{array}{@{}ll}
\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
            \begin{array}{l}
                \pair{[]}{\_}{} 
                \\\\
                \bigRedBracket{
                \begin{array}{l} 
                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
                \end{array}}
            \end{array}}
\end{array}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[} 
                                          \DVar{\!},
                                          \DConst{\!},
                                          \DProd{\!}{\!},
                                          \DSigma{\!}{\!}, 
                                          \DPi{\!}{\!}
                                             \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

@@

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}
\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
                          \end{array}
                          }
\end{array}
}\]

Intuitively, the constraint is captured as follows. First, we store an
element $\V{m}$ of $\Nat$, thanks to a $\DSigma{\!}{\!}$. However, the
constraint stipulates that $\V{m}$ cannot be \emph{any} natural
numbers: it must be ``the index minus one''. This translates into the
constraint $\V{n} \PropEq \NatSuc{\V{m}}$.

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PITEL{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
    \pair{\pair{\ListCons{\DVar{\!}}{(\fst{\V{E}})}} 
               {\LAM{\V{i}}
                \pair{(\DConst{(\V{R}\: \V{i})})}
                     {((\snd{\V{E}})\: \V{i})}{}}{}}
         {\V{F}}{}
}\]

@@

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~\V{I} \mapsto \DSigma{\!}{\!} & (\EnumT \red{[}
                                              \DConst{\!},
                                              \DProd{\!}{\!},
                                              \DSigma{\!}{\!}, 
                                              \DPi{\!}{\!}
                                                    \red{]}) \\
                                  & \bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}
\end{array}
}\]

@@
